---
output: github_document
---
<!-- README.md is generated from README.Rmd. Please edit that file -->
<!-- cSpell: words Makevars -->

```{r, include = FALSE}
knitr::opts_chunk$set(
    collapse = TRUE,
    comment = "#>",
    fig.path = "man/figures/README-",
    out.width = "100%"
)
```


# integratecpp

<!-- badges: start -->
<!-- markdownlint-capture -->
<!-- markdownlint-disable MD045 -->
[![Codecov test coverage](https://codecov.io/gh/hsloot/integratecpp/branch/main/graph/badge.svg)](https://app.codecov.io/gh/hsloot/integratecpp?branch=main)
[![](https://img.shields.io/badge/pkgdown-integratecpp-blue.svg)](https://hsloot.github.io/integratecpp/)
[![](https://img.shields.io/badge/Doxygen-integratecpp-blue.svg)](https://hsloot.github.io/integratecpp/html/index.html)
<!-- markdownlint-restore -->
<!-- badges: end -->

The package `integratecpp` provides a header-only C++11 interface to R's C-API
for numerical integration.


## Installation

You can install the development version of `integratecpp` like so:

```{r, eval=FALSE}
# R
remotes::install_github("hsloot/integratecpp")
```

To include the header into your C++ source files for building with Rcpp, use

```{Rcpp, eval=FALSE}
// C++
// [[Rcpp::plugins(cpp11)]]
// [[Rcpp::depends(integratecpp)]]

#include <integratecpp.h>
// your code
```

To use the header in source files of an R-package, include the following lines
in your `DESCRIPTION` file:

```deb-control
LinkingTo: integratecpp
```

Note that the header includes only `C++` standard library headers and
[`<R_ext/Applic.h>`](https://github.com/wch/r-source/blob/trunk/src/include/R_ext/Applic.h).


## How to use it?

Suppose you want to integrate the identity function over the unit interval. 
First, you include the required headers (and Rcpp attributes, if Rcpp is used),
for example:

```{Rcpp, eval=FALSE}
#include <integratecpp.h>
```

Second, you have to define the integrand as a `Callable` that is invocable by
`const double` and returns `double`, for example a lambda-functor:

```{Rcpp, eval=FALSE}
auto fn = [](const double x) {
    return x;
};
```

Third, you can use the integrate routine:

```{Rcpp, eval=FALSE}
const auto result = integratecpp::integrate(fn, 0., 1.);
```

Better, the last part should be enclosed in a try-catch block:

```{Rcpp, eval=FALSE}
try {
  const auto result = integratecpp::integrate(fn, 0., 1.);
  // ...
} catch (const integratecpp::integration_logic_error &e) {
  // ...
} catch (const integratecpp::integration_runtime_error &e) {
  // ...
}
```

Optional configurations similar to those of `stats::integrate` are also
available if needed. For a more realistic example, see 
["Using `integratecpp`"](https://hsloot.github.io/integratecpp/articles/integratecpp.html).


## Why is this useful?

Many R package authors implement critical parts, including numeric
integration, in C, Fortran or C++ to improve performance. However, while R
provides an
[API for C](https://cran.r-project.org/doc/manuals/r-release/R-exts.html#The-R-API) and
it is possible to [mix C and C++](https://isocpp.org/wiki/faq/mixing-c-and-cpp),
using the C-API in C++ code can pose a higher burden for those more familiar
with R and Rcpp than C++ or C.

Using the C-API for the numerical integration routines requires adhering to
the interface of the functions `Rdqags` or `Rdqagi`. This can be done by
creating a callback functor taking a `void *` pointer to the original function,
which is then internally cast to the correct type and is used to overwrite an
array of doubles with corresponding function evaluations. This is rather
complicated and requires being more familiarity with pointers. Additionally, it
requires translating error codes into a proper error message. To make it worse,
not guarding callback functions against C++ exceptions introduces possible
undefined behavior.

This packages bridges this gap by providing a simple, easy-to-use C++11 wrapper
for R's C-API for numerical integration.


## Alternatives

There are alternatives to using `integratecpp` or R's C-API for numerical
integration in compiled code of R packages. Two examples are:

- Linking to the [GNU scientific library](https://www.gnu.org/software/gsl/),
  possibly using [RcppGSL](https://github.com/eddelbuettel/rcppgsl).
- Using [RcppNumerical](https://github.com/yixuan/RcppNumerical), which
  provides a similar approach to ours.

Both approaches provide a finer control over the specific integration algorithms
than R does. The following table provides a summary.

<!-- markdownlint-capture -->
<!-- markdownlint-disable MD013 -->

| __Approach__   | __Depends__ | __Imports__ | __LinkingTo__   | __SystemRequirements__ | __External dependency__ | __Additional features__ |
|:---------------|-------------|-------------|-----------------|------------------------|-------------------------|-------------------------|
| `integratecpp` | `R >= 3.1`  | (Rcpp[^1])  | (Rcpp)          | C++11                  |                         | &#10060;                |
| C-API          |             |             |                 |                        |                         | &#10060;                |
| gsl            |             | (Rcpp)      | (Rcpp, RcppGSL) |                        | gsl                     | &#9989;                 |
| RcppNumerical  |             | Rcpp        | Rcpp, RcppEigen |                        |                         | &#9989;                 |

<!--markdownlint-restore -->

What separates our approach are little additional dependencies (zero, if
vendored) and an intuitive pure-C++ API which does not rely on Rcpp itself.
Hence, as `Rdqags` and `Rdqagi` are not using longjumps *themselves*, our
approach can be used in a pure C++ back-end.[^2] A comparison of different
numerical integration approaches in C++ is summarized in the article
["Comparing numerical integration packages"](https://hsloot.github.io/integratecpp/articles/web_only/comparison.html).


## Outlook

The current version of `integratecpp` has the following shortcomings which could
be addressed in future versions:

- We currently import and link to Rcpp to generate test functions which are not
  exported. Future versions might remove this dependency, see
  [#8](https://github.com/hsloot/integratecpp/issues/8).
- R's C-API for numerical integration allows reusing workspace variables. We
  have not implemented this feature, i.e., each call to
  `integratecpp::integrate(...)` or `integratecpp::integrator::operator()(...)`
  will create a `std::vector<int>` and a `std::vector<double>` of length `limit`
  and `4 * limit`, respectively. Future versions might make workspace variables
  class members of the `integratecpp::integrator` or allow to provide them
  externally, see
  [#9](https://github.com/hsloot/integratecpp/issues/9).
- The current version of `integratecpp` is licensed under `GPL (>=3)` due to its
  dependence and linking to Rcpp, the header-only library is licensed under
  `LGPL (>=3)` as it only depends only on STL headers and
  [`<R_ext/Applic.h>`](https://github.com/wch/r-source/blob/trunk/src/include/R_ext/Applic.h).
  The latter is licensed under `LGPL (>= 2.1)` (see
  [`doc/COPYRIGHTS`](https://github.com/wch/r-source/blob/trunk/doc/COPYRIGHTS)).
  Hence, future versions might relicense the package under a more permissive
  library if the Rcpp dependency can be removed, see
  [#10](https://github.com/hsloot/integratecpp/issues/10).


## Code of Conduct

Please note that the `integratecpp` project is released with a
[Contributor Code of Conduct](.github/CODE_OF_CONDUCT.md). By contributing to
this project, you agree to abide by its terms.

[^1]: The current version of `integratecpp` imports and links to Rcpp for
internal testing. This dependency might be removed in future versions and can be
avoided if the header is vendored into the project.

[^2]: Note that the provided integrator-function itself should not use longjumps
(e.g., R functions provided through Rcpp) to exit the frame; if you want to use
R functions with `Rcpp >= 1.0.10`, consider defining `RCPP_NO_UNWIND_PROTECT`
before loading the Rcpp header file.
