---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# integratecpp

<!-- badges: start -->
<!-- badges: end -->

The package `integratecpp` provides a header-only interface to `R`'s numerical integration methods.

## Installation

You can install the development version of integratecpp like so:

``` r
remotes::install_github("hsloot/integratecpp")
```

## Example

This is a basic example which shows you how to solve a common problem:
We want to calculate the expectation of an exponential distribution with rate $\lambda$ by numerical integration.
The theoretical expectation is
$$
  \int_{0}^{\infty}{ x \lambda \exp{\{ -\lambda x \}} } \mathrm{d}x
    = \frac{1}{\lambda} .
$$

```{Rcpp example}
// C++

// [[Rcpp::plugins(cpp11)]]
// [[Rcpp::depends(integratecpp)]]

#include <cmath>
#include <limits>
#include <stdexcept>

#include <Rcpp.h>

#include "integratecpp.h"

// [[Rcpp::export(rng=false)]]
Rcpp::List integrate_exponential_expectation(const double lambda) {
  auto fn = [lambda](const double x) {
    return x * lambda * std::exp(-lambda * x);
  };

  decltype(integratecpp::integrate(fn, 0., 1.)) result;
  try {
    result =
        integratecpp::integrate(fn, 0., std::numeric_limits<double>::infinity());
  } catch (const integratecpp::integration_runtime_error &e) {
    result = e.result();
  } catch (const integratecpp::integration_logic_error &e) {
    result = e.result();
  } catch (const std::exception &e) {
    Rcpp::stop(e.what());
  } catch (...) {
    Rcpp::stop("Unexcpected error");
  }

  return Rcpp::List::create(Rcpp::Named("value") = result.value,
                            Rcpp::Named("abserr") = result.abserr,
                            Rcpp::Named("subdivisions") = result.subdivisions,
                            Rcpp::Named("neval") = result.neval);
}
```


```{r}
## R

integrate_exponential_expectation(1)
integrate_exponential_expectation(2)
integrate_exponential_expectation(3)

integrate_exponential_expectation(-1)
```


## Code of Conduct

Please note that the integratecpp project is released with a [Contributor Code of Conduct](https://contributor-covenant.org/version/2/0/CODE_OF_CONDUCT.html). By contributing to this project, you agree to abide by its terms.
