---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# integratecpp

<!-- badges: start -->
[![Codecov test coverage](https://codecov.io/gh/hsloot/integratecpp/branch/main/graph/badge.svg)](https://app.codecov.io/gh/hsloot/integratecpp?branch=main)
<!-- badges: end -->

The package `integratecpp` provides a header-only `C++11` interface to `R`'s `C`-API for numerical integration.

## Installation

You can install the development version of `integratecpp` like so:

``` r
remotes::install_github("hsloot/integratecpp")
```

To include the header into your `C++` source files for building with `Rcpp`, use

```{Rcpp, eval=FALSE}
// C++

// [[Rcpp::plugins(cpp11)]]
// [[Rcpp::depends(integratecpp)]]

#include <integratecpp.h>

// your code
```

If you intend including the header in source files of an `R`-package, you should drop the `Rcpp`
attributes and include the following line in your `DESCRIPTION` file:

```batch
LinkingTo: integratecpp
```

Note that the header does includes only C++ standard library headers and [`<R_ext/Applic.h>`](https://github.com/wch/r-source/blob/trunk/src/include/R_ext/Applic.h).


## Why is this useful?

Many `R` package authors implement critical parts in `C`, `Fortran` or `C++` to improve performance.
However, while `R` provides an [API for `C`](https://cran.r-project.org/doc/manuals/r-release/R-exts.html#The-R-API)
and it is possible to [mix `C` and `C++`](https://isocpp.org/wiki/faq/mixing-c-and-cpp), using the
`C`-API in `C++` code can pose a higher burden for those more familiar with `R` and `Rcpp` than
`C++` or `C`.

Consider the following example, approximating the integral of the identity function $x \mapsto x$ over the interval $[0, 1]$:

```{Rcpp}
// C++

// [[Rcpp::plugins(cpp11)]]
// [[Rcpp::depends(integratecpp)]]

#include <algorithm> // std::transform

#include <Rcpp.h>
#include <R_ext/Applic.h>

// [[Rcpp::export(rng=false)]]
Rcpp::List integrate_identity() {
  // define integrand as lambda
  auto fn = [](const double x) {
    return x;
  };

  // define bounds
  double lower = 0.;
  double upper = 1.;

  // initialize output variables for integration results
  double result;
  double abserr;
  int last;
  int neval;

  // initialize configuration parameters
  int limit = 100;
  double epsrel = 0.0001220703125;
  double epsabs = epsrel;
  int lenw = 4 * limit;

  // initialize working array
  auto iwork = std::vector<int>(limit);
  auto work = std::vector<double>(lenw);

  // initialize variable for error code
  int ier = 0;

  // NOTE: `Rdqags` requires a function pointer with signature
  // `void(*)(double *, int, void *)` and a void pointer
  // `void *` passed to the callback as the last argument
  const auto fn_callback = [](double *x, int n, void *ex) {
    auto& fn_integrand = *static_cast<decltype(&fn)>(ex);
    std::transform(&x[0], &x[n], &x[0], fn_integrand);
    return;
  };

  // NOTE: finite bounds can be integrated with the `C`-method `Rdqags`
  Rdqags(fn_callback, &fn, &lower, &upper, &epsabs, &epsrel, &result,
         &abserr, &neval, &ier, &limit, &lenw, &last, iwork.data(),
         work.data());

  return Rcpp::List::create(
      Rcpp::Named("value") = result,
      Rcpp::Named("absolute_error") = abserr,
      Rcpp::Named("subdivisions") = last,
      Rcpp::Named("neval") = neval);
}
```

```{r}
integrate_identity()
```

The key part for adhering to the interface of `Rdqags` is creating the callback functor
`fn_callback`, taking a `void *` pointer to the original function, which is than internally casted
to the correct type and is used to overwrite an array of double with corresponding function evaluations.
This is rather complicated and requires being more familiarity with pointers.
Additionally, this snippet is missing a translation of the error code into a proper error message.
To make it worse, without guarding the callback functions from `C++` exceptions, we introduce
possible undefined behavior.

Using the wrapper simplifies numerical integration in `Rcpp` considerably:

```{Rcpp}
// C++

// [[Rcpp::plugins(cpp11)]]
// [[Rcpp::depends(integratecpp)]]

#include <stdexcept>
#include <exception>

#include <Rcpp.h>
#include <integratecpp.h>

// [[Rcpp::export(rng=false)]]
Rcpp::List integrate_identity_improved() {
  auto fn = [](const double x) {
    return x;
  };

  try {
    const auto result = integratecpp::integrate(fn, 0., 1.);
    return Rcpp::List::create(
        Rcpp::Named("value") = result.value,
        Rcpp::Named("absolute_error") = result.absolute_error,
        Rcpp::Named("subdivisions") = result.subdivisions,
        Rcpp::Named("neval") = result.neval);
  } catch (const std::exception& e) {
    Rcpp::stop(e.what());
  }
}

// [[Rcpp::export(rng=false)]]
Rcpp::List integrate_identity_error() {
  auto fn = [](const double x) {
    throw std::runtime_error("stop on purpose");
    return x;
  };

  try {
    const auto result = integratecpp::integrate(fn, 0., 1.);
    return Rcpp::List::create(
        Rcpp::Named("value") = result.value,
        Rcpp::Named("absolute_error") = result.absolute_error,
        Rcpp::Named("subdivisions") = result.subdivisions,
        Rcpp::Named("neval") = result.neval);
  } catch (const std::exception& e) {
    Rcpp::stop(e.what());
  }
}
```

```{r}
integrate_identity_improved()
tryCatch(integrate_identity_error(), error = function(cond) print(cond))
```

## Code of Conduct

Please note that the integratecpp project is released with a [Contributor Code of Conduct](https://contributor-covenant.org/version/2/0/CODE_OF_CONDUCT.html). By contributing to this project, you agree to abide by its terms.
