---
title: "Using `integratecpp`"
output:
  rmarkdown::html_vignette:
    code_folding: hide
vignette: >
  %\VignetteIndexEntry{Using `integratecpp`}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---
<!-- cSpell: ignoreRegExp \\.* -->
<!-- cSpell: words cmath,stdexcept -->
<!-- cSpell: words decltype,nullptr,pgamma,isfinite -->
<!-- cSpell: words cond,abserr,neval,epsrel,epsabs,lenw,iwork -->

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  class.source = "fold-none"
)
```

The following vignette demonstrates how using `integratecpp` simplifies and improves using numerical
integration in `Rcpp`. [^1]
We use a somewhat simplified problem to hint possible use-cases for `integratecpp` and demonstrate
its usage.
Suppose, we want need to evaluate the [*upper incomplete Gamma function*](https://en.wikipedia.org/wiki/Incomplete_gamma_function)
$$
    \Gamma{(\alpha, x)}
        = \int_{x}^{\infty}{
            \exp{\{ - u \}}  u^{\alpha-1}
            {\mathrm{d}u}
        } ,
        \quad x > 0 ,
$$
for values $\alpha \in \mathbb{R}$, $x \geq 0$.[^2][^3]

# An implementation with `integratecpp`

```{Rcpp run-all-C++,eval=TRUE, echo=FALSE, ref.label=knitr::all_rcpp_labels(), class.source = 'fold-hide'}
```

As `integratecpp` requires C++11, we use the `Rcpp` plugin `cpp11` and link to the
includes directory of `integratecpp` using the depends attribute `integratecpp`.

```{Rcpp attributes, eval=FALSE, echo=TRUE, class.source = 'fold-none'}
// C++11

// [[Rcpp::plugins(cpp11)]]
// [[Rcpp::depends(integratecpp)]]
```

Next, we include the STL header `<cmath>` for using `std::pow` and we include the header
`<integratecpp.h>`.

```{Rcpp includes, echo=TRUE, eval=FALSE}
#include <cmath>
#include <integratecpp.h>
```

Finally, we implement our function using the `integratecpp` interface to
`Rdqagi`:

```{Rcpp demo-Rcpp, echo=TRUE, eval=FALSE}
// [[Rcpp::export]]
double gamma_inc(const double alpha, const double x) {
    // NOTE: integrand implemented as capturing lambda
    const auto fn = [alpha](const double u) {
        return std::exp(-u) * std::pow(u, alpha - 1.);
    };

    // NOTE: throws if integration error occurs
    const auto result = integratecpp::integrate(
        fn, x, std::numeric_limits<double>::infinity());

    return result.value;
}
```

```{r demo-r}
# R

## demo with sane input
gamma_inc(-0.5, 1)

## demo with not so sane input
tryCatch(gamma_inc(0, 0), error = function(cond) print(cond))
tryCatch(gamma_inc(0, -1), error = function(cond) print(cond))
```

Note, that in this basic form, using `integratecpp`'s wrapper requires only the
following two things:

- A [`Callable`](https://en.cppreference.com/w/cpp/named_req/Callable)
  object which is [invocable](https://en.cppreference.com/w/cpp/types/is_invocable)
  with type `const double` and yields a result convertible to `double`.
  Roughly, an object that can be used to construct `std::function<double(const double)>`,
  e.g., a suitable
  [lambda expression](https://en.cppreference.com/w/cpp/language/lambda).
- A call to `integratecpp::integrate`.

Optionally, integration parameters can be changed,
exceptions can be captured and handled,
and the integration error estimate can be further examined.

# An implementation without `integratecpp`

The following example does not use the `integratecpp` wrapper, but hints how the wrapper works
internally.
First, we need to include the header for R's C entry-points for the integration routines `Rdqags` 
and `Rdqagi`:

```{Rcpp included-additional-1, echo=TRUE, eval=FALSE}
// C++11

#include <R_ext/Applic.h>
```

Then, we require some STL headers for handling data:

```{Rcpp includes-additional-2, echo=TRUE, eval=FALSE}
#include <algorithm> // std::transform
#include <utility>   // std::make_pair, std::pair
#include <vector>    // std::vector
```

Also, we need more STL headers and the `Rcpp` header for exception and error handling:

```{Rcpp includes-additional-3, echo=TRUE, eval=FALSE}
#include <exception>  // std::exception_ptr, std::make_exception_ptr, 
                      // std::rethrow_exception, std::current_exception
#include <stdexcept>  // std::exception, std::runtime_error
#include <Rcpp.h>     // Rcpp::stop
```

Finally, we can implement the *upper incomplete Gamma function*:

```{Rcpp demo-Rcpp-direct, echo=TRUE, eval=FALSE}
// [[Rcpp::export]]
double gamma_inc_direct(const double alpha, const double x) {
    // NOTE: integrand implemented as capturing lambda
    const auto fn = [alpha](const double u) {
        return std::exp(-u) * std::pow(u, alpha - 1.);
    };

    // create bounds parameters for `Rdqagi`
    double bound = x;
    int inf = 1.;

    // initialize output variables for integration results
    double result;
    double abserr;
    int last;
    int neval;

    // initialize configuration parameters
    int limit = 100;
    double epsrel = 0.0001220703125;
    double epsabs = epsrel;
    int lenw = 4 * limit;

    // initialize working array
    auto iwork = std::vector<int>(limit);
    auto work = std::vector<double>(lenw);

    // initialize variable for error code
    int ier = 0;

    // NOTE: `Rdqagi` requires a function pointer with signature
    // `void(*)(double *, int, void *)` and a void pointer
    // `void *` passed to the callback as the last argument
    const auto fn_callback = [](double *x, int n, void *ex) {
        // cast `void *` to original type and create references
        // to the `Callable` and the exception pointer
        using ex_t = std::pair<decltype(fn), std::exception_ptr>;
        auto &fn_integrand = (*static_cast<ex_t *>(ex)).first;
        auto &e_ptr = (*static_cast<ex_t *>(ex)).second;
        
        // evaluate `Callable` at *in-out* pointer-to-array `x`
        // and catch possible exceptions in exception pointer
        try {
            std::transform(&x[0], &x[n], &x[0], fn_integrand);
        } catch (const std::exception& e) {
            e_ptr = std::current_exception();
        }
        
        // store a runtime exception in the exception pointer 
        // if any results are infinite 
        if (!std::all_of(&x[0], &x[n], 
                         [](const double x) { 
                             return std::isfinite(x); 
                         })) {
            e_ptr = std::make_exception_ptr(
                std::runtime_error("non-finite function value"));
          }
    };

    // create pair with `Callable` and exception pointer 
    // (`nullptr` on initialization) as external data for 
    // callback function
    auto ex = std::make_pair(fn, std::exception_ptr());
    
    // call C-method `Rdqagi`
    Rdqagi(fn_callback, &ex, &bound, &inf, &epsabs, &epsrel, &result,
           &abserr, &neval, &ier, &limit, &lenw, &last, iwork.data(),
           work.data());
    
    // rethrow possible exceptions during function evaluation
    if (ex.second) std::rethrow_exception(ex.second);
    
    // throw `Rcpp::exception` if integration error occurred
    if (ier > 0) Rcpp::stop("Integration error");

    return result;
}
```

```{r demo-r-direct}
# R

## demo with sane input
gamma_inc_direct(-0.5, 1)

## demo with not so sane input
tryCatch(gamma_inc_direct(0, 0), error = function(cond) print(cond))
tryCatch(gamma_inc_direct(0, -1), error = function(cond) print(cond))
```


[^1]: We use `Rcpp` for simplicity. There no inherent requirement to use `Rcpp` with `integratecpp`.
Using `cpp11` or writing your own wrapper is also possible.

[^2]: The *upper incomplete Gamma function* can be indirectly evaluated for non-negative $\alpha$
with the `pgamma`-function, which even is provided as a
[entry-point in Rs C-API](https://cran.r-project.org/doc/manuals/r-release/R-exts.html#Distribution-functions).
In contrast, this is not possible for negative $\alpha$ as of `R 4.2-0`.

[^3]: The package `gsl` provides a function and entry-points for C code to evaluate the
*upper incomplete Gamma function* for negative values.
