---
title: "Comparing numerical integration packages"
references:
- id: Piessens1983a
  title: Quadpack
  author:
  - given: Robert
    family: Piessens
  - given: Elise
    family: Doncker-Kapenga
  - given: Christoph W.
    family: Überhuber
  - given: David K.
    family: Kahaner
  DOI: 10.1007/978-3-642-61786-7
  publisher: Springer Berlin, Heidelberg
  type: book
  issued:
    year: 1983
nocite: |
  @Piessens1983a
---

```{r knitr-setup, include = FALSE}
knitr::opts_chunk$set(
    collapse = TRUE,
    comment = "#>",
    cache = TRUE
)
```

## The Problem

We are using the following simple integral to compare various numerical integration options:

$$
    \int_{0}^{1} 1 / x^{0.7} \mathrm{d}x .
$$


## Using base `R`

First, we calculate the integral using base-`R`.

- Finite-interval integration uses a 21-point *Gauss-Kronrod* rule on equally-spaced subintervals.
- Infinite-interval integration uses a 15-point *Gauss-Kronrod* rule on equally-spaced subintervals after transforming the integrand to the interval $[0, 1]$.
- Both use *Wynn's $\epsilon$ algorithm* for limit-extrapolation.

It is possible to configure the integration algorithms by changing the requested relative and absolute accuracies and the maximum number of subintervals.
All of these features are shared by the following approaches.
However, some have additional features.

```{r base-R}
# R

integrate_R <- function() {
    integrate(
        function(x) 1 / x^0.7,
        lower = 0, upper = 1
    ) |>
    (function(x) {
        list(
            "value" = x$value,
            "abserr" = x$abs.err,
            "subdivisions" = x$subdivisions)
    })()
}

integrate_R()
```


## Using `R`'s `C`-API

We can calculate the same integral using `R`'s `C`-API

```{Rcpp C-API}
// C++11 

#include <math.h>

#include <R_ext/Applic.h>
#include <Rcpp.h>

void fn(double *x, int n, void* ex) {
    for (int i = 0; i < n; ++i) {
        x[i] = 1. / pow(x[i], 0.7);
    }
    return;
}

// [[Rcpp::export(rng=false)]]
Rcpp::List integrate_Cish() {
    double lower = 0.;
    double upper = 1.;

    double result;
    double abserr;
    int last;
    int neval;

    int limit = 100;
    double epsrel = 0.0001220703125;
    double epsabs = epsrel;
    int lenw = 4 * limit;

    #pragma GCC diagnostic push
    #pragma GCC diagnostic ignored "-Wvla-extension"
    int iwork[limit];  // C99
    double work[lenw]; // C99
    #pragma GCC diagnostic pop

    int ier = 0;

    Rdqags(&fn, nullptr, &lower, &upper, &epsabs, &epsrel, &result,
           &abserr, &neval, &ier, &limit, &lenw, &last, iwork, work
    );

    return Rcpp::List::create(
        Rcpp::Named("value") = result,
        Rcpp::Named("abserr") = abserr,
        Rcpp::Named("subdivisions") = last
    );
}
```

```{r C-API-eval}
# R

integrate_Cish()
```


## Using `integratecpp`

Using our `C++11` wrapper, numerical integration can be done seamlessly in `Rcpp`:

```{Rcpp integratecpp}
// C++11 

// [[Rcpp::plugins(cpp11)]]
// [[Rcpp::depends(integratecpp)]]

#include <cmath>

#include <Rcpp.h>
#include <integratecpp.h>

// [[Rcpp::export(rng=false)]]
Rcpp::List integrate_integratecpp() {
    const auto fn = [](const double x) {
        return 1. / std::pow(x, 0.7);
    };

    const auto result = integratecpp::integrate(fn, 0., 1.);

    return Rcpp::List::create(
        Rcpp::Named("value") = result.value,
        Rcpp::Named("abserr") = result.absolute_error,
        Rcpp::Named("subdivisions") = result.subdivisions
    );
}
```

```{r integratecpp-eval}
# R

integrate_integratecpp()
```


## Using `RcppGSL`

With `RcppGSL`, we can use `gsl` in a similar way as `R`s `C`-API:

```{Rcpp RcppGSL}
// C++11 

// [[Rcpp::depends(RcppGSL)]]

#include <math.h>

#include <Rcpp.h>
#include <RcppGSL.h>
#include <gsl/gsl_integration.h>

double fn(double x, void* params) {
    return 1. / pow(x, 0.7);
}

// [[Rcpp::export(rng=false)]]
Rcpp::List integrate_RcppGSL() {
    double lower = 0.;
    double upper = 1.;

    double result;
    double abserr;

    size_t limit = 100;
    double epsrel = 0.0001220703125;
    double epsabs = epsrel;

    gsl_function f;
    f.function = &fn;
    f.params = nullptr;

    gsl_integration_workspace *w =
      gsl_integration_workspace_alloc(limit);

    gsl_integration_qags(
        &f, lower, upper, epsabs, epsrel, limit, w, &result, &abserr
    );
    int subintervals = w->size;

    gsl_integration_workspace_free(w);

    return Rcpp::List::create(
        Rcpp::Named("value") = result,
        Rcpp::Named("abserr") = abserr,
        Rcpp::Named("subdivisions") = subintervals
    );
}
```

```{r RcppGSL-eval}
# R

integrate_RcppGSL()
```


## Using `RcppNumerical`

With `RcppNumerical`, we must define the integrator in the `Numer` namespace:

```{Rcpp RcppNumerical}
// C++11

// [[Rcpp::depends(RcppEigen)]]
// [[Rcpp::depends(RcppNumerical)]]

#include <cmath>

#include <RcppNumerical.h>

namespace Numer {

class Fn : public Func {
    double operator()(const double& x) const {
        return 1. / std::pow(x, 0.7);
    }

};

} // Numer

// [[Rcpp::export]]
Rcpp::List integrate_RcppNumerical() {
    double lower = 0.;
    double upper = 1.;

    size_t limit = 100;
    double epsrel = 0.0001220703125;
    double epsabs = epsrel;

    Numer::Fn fn;

    double abserr;
    int ier;

    const double res = Numer::integrate(
        fn, lower, upper, abserr, ier, limit, epsabs, epsrel,
        Numer::Integrator<double>::GaussKronrod21
    );

    return Rcpp::List::create(
        Rcpp::Named("value") = res,
        Rcpp::Named("abserr") = abserr
    );
}
```

```{r RcppNumerical-eval}
# R

integrate_RcppNumerical()
```


## Benchmark

A benchmarks shows that our wrapper is almost as fast as the native `C`-API.

```{r bench}
# R

bench::mark(
    integrate_R(),
    integrate_integratecpp(),
    integrate_Cish(),
    integrate_RcppGSL(),
    integrate_RcppNumerical(),
    check = FALSE
)
```


## References
